Evolution of ORM Systems: Formal Foundations

### 2013-01-24 doc/description.tex
formating

\section{ORM Software}
\label{sec:evoIntro}
\begin{figure}
\centering
	\includegraphics[scale=0.3]{./images/system}
	\caption{The model of one generation of an ORM system consists of application persistent layer and relational database for storing instances.}
\label{fig:appStructure}
\end{figure}
A software implemented using an object-oriented language, which uses relational database as a data storage (as illustrated in the Figure \ref{fig:appStructure}) is object of interest of this paper. There are three important components of such a software: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM (we will use $\rho$ as a symbol for ORM), therefore we define a software as a triple:
\begin{align*}
& \mathbf{software} = < Application, Database, \rho >\\
& \mathbf{sw} \in Applications \times Databases \times ORM \rightarrow  Software \\ \\
& \mathbf{application} \in Software \rightarrow Application \\
& application(sw(a, d, \rho)) = a \\ \\
& \mathbf{database} \in Software \rightarrow Database \\
& database(sw(a, d, \rho)) = d \\\\
& a \in Application, d \in Database
\end{align*}

@@

The function querying the state (it means consistency) of a software is defined as:
\begin{align*}
& 	\mathbf{consistent} \in State \rightarrow Boolean \\
& 	consistent(s) = \begin{cases}
 		\inference{s = Consistent(s)}{true}\\\\
 		\inference{s = Inconsistent(s)}{false}\\
 \end{cases}\\
& s \in State
\end{align*}
The consistency of application and database is discussed later in this paper. The definition of software state only is provided:
\begin{align*}
&	\mathbf{state} \in Software \rightarrow State \\
&	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \\ \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases} \\ 
& 	s \in State
\end{align*}

@@

The application of a transformation on a software state (evolution) is marked by the $\oplus$ operator:
\begin{align*}
& \oplus \in State \times E \rightarrow State
\end{align*}
and similarly the $\circ$ and the $\bullet$ operators for evolving application and database:
\begin{align*}
& \circ \in State \times AppT \rightarrow State \\
& \bullet \in State \times DbT \rightarrow State
\end{align*}
The definition of the evolving operator is inspired by the monads (as used e.g. in Haskell programming language) so the state of the software can be passed from one evolution step to another, which helps us to concatenate evolution transformations.
\begin{align*}
& state(s) \oplus t = \begin{cases}
 	\inference{! consistent(state(s))}{Inconsistent s} \\ \\
 	\inference{consistent(state(s))}{state(sw(\Psi(s, t), \Phi(s,t))} 
\end{cases}\\ 
& s \in State, t \in E
\end{align*}
\begin{align*}
& state(a) \circ e = \begin{cases}
 	\inference{!\;consistent(state(a))}{a} \\ \\
 	\inference{consistent(state(a))}{e(a)} 
\end{cases}\\ 
& a \in Application, e \in AppT
\end{align*}
\begin{align*}
&  state(d) \bullet e = \begin{cases}
 	\inference{!\;consistent(state(d))}{d} \\ \\
 	\inference{consistent(state(d))}{e(d)} 
\end{cases}\\ 
& d \in Database, e \in DbT
\end{align*}

@@

The definitions of all parts assume a set $Label$ exists which contains all possible identifiers and each modeled entity is unambiguously identified in the model. Next there is a set $Values$ containing all possible values, which can be used in the model. The notation: $Type*$ is used to define an ordered set of values of concrete type (e.g. $C \in Class*$ means c is an ordered set of classes).

A lot of introduced transformations can be applied on an ordered set. The application proceeds on set consists in sequential application on set elements:
\begin{align*}
& 	f(x, y) = \begin{cases}
 		\inference{x = \varnothing}{y}\\
 		\inference{|x| > 1}{f(head(x), y) \cup f(tail(x), y}\\\\
 		\inference{|x| = 1}{effect}\\\\
 		\inference{otherwise}{Inconsistent y}
 \end{cases}\\\\
&  y \in Software \vee y \in Application \vee y \in Database
\end{align*}
where $x$ is a set of parameters. The effect of a transformation is provided only in next sections as the repetition of sequential application will be redundant. The $otherwise$ branch is consider implicit for all introduced transformations and thus it is not part of transformation definitions too.

@@

\paragraph{Foreign key} Foreign key is a reference to another table's primary key, it has an unique name and it can be constrained. The value of a foreign key is $\varnothing$ or a non-zero natural number.
\begin{align*}
&	\mathbf{ForeignKey} = <label, tableRef, Constraint*> \\ 
&	\mathbf{fk} : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
&	\mathbf{name} : ForeignKey \rightarrow Label \\
&	name(fk(l, tRef, cons)) = lab  \\ \\
&	\mathbf{reference} : ForeignKey \rightarrow Label  \\
&	reference(fk(l, tRef, cons)) = tRef  \\ \\
&	\mathbf{constaints} : ForeignKey \rightarrow Constraint*  \\
&	constaints(fk(l, tRef, cons)) = cons  \\ \\
&	\mathbf{owningTable} : ForeignKey \times Database \rightarrow Table  \\
&	owningTable(fk, d) = \inference{\exists tab \in tables(d) \wedge fk \in foreignKeys(tab)}{tab}\\\\
&	l, tRef \in Label, cons  = \{ x | x \in Constraint\}, \\ 
& d \in Database, fk = \{ x | x \in ForeignKey\}
\end{align*}

@@

\paragraph{Stored Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete value of concrete column. Each row contains a name of a table it belongs to.
\begin{align*}
&	\mathbf{Row} = < refTable, pkValue, pair* > \\
&	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
&	\mathbf{refTable} : Row \rightarrow Label \\
&	refTable(row(l, n, ps) = l \\ \\
&	\mathbf{pairs} : Row \rightarrow Pair* \\
&	pairs(row(l, n, ps)) = ps \\ \\
&	\mathbf{pkValue} : Row \rightarrow N \\
&	pkValue(row(l, n, ps)) = n \\ \\
&	 l \in Label, ps = \{ x | x \in Pair\}, n \in N
\end{align*}
The value stored in concrete column of the row is represented by a pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
&	\mathbf{Pair} = < refCol, value > \\
&	[Label \times Value] \rightarrow Pair\\ \\
&	\mathbf{reference} : Pair \rightarrow Label \\
&	reference([l,val]) = l \\ \\
&	\mathbf{value} : Pair \rightarrow Value \\ 
&	value([l,val]) = val \\ \\
&	l \in Label, val \in Value
\end{align*}
The definition of data allows definition of function $contains$ which verify if the table contains data in context of concrete database and $instantiated$, which search for instances of a class with given name and returns true if there are rows for the table in the database.
\begin{align*}
&	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
&	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
& l \in Label, ps  = \{ x | x \in Pair\}, t \in tables(d), d \in Database
\end{align*}
\begin{align*}
&	\mathbf{tableData} : Label \times Row* \rightarrow Row* \\
&	tableData(l, r) = \begin{cases}
 		%\inference{r = \varnothing}{\varnothing} \\
 		%\inference{|r| > 1}{tableData(head(r)) \cup tableData(tail(r))}\\ \\
&		\inference{|r| = 1 \wedge refTable(r) = l}{r}\\\\
& 		\inference{|r| = 1 \wedge refTable(r) \neq l}{\varnothing}
 \end{cases} \\ \\
&	r \in Row*, l \in Label
\end{align*}
\begin{align*}
&	\mathbf{instantiated} : Class \times Software \rightarrow Boolean \\
&	instantiated(c, s) = \begin{cases} \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  false
 \end{cases} \\ \\
&	 c \in classes(application(s)), s \in State 
\end{align*}

\subsubsection{Database consistency}
The database consistency is founded on the consistency of structure and consistency between structure and stored values. The conditions for structure consistency are similar as for application structure. The consistency between structure and data is:  each row in a database  refers to a table in database schema and it has to have the structure corresponding to the one defined by a table, moreover the column and foreign key constrains has to be fulfilled  otherwise the database is inconsistent. The consistency of database is:
\begin{align*}
&	\mathbf{consistent} \in Database \rightarrow Boolean \\
&	state(d) = \begin{cases}
&		\inference{ \forall t_1, t_2 \in tables(d) : name(t_1) \neq name(t_2) \\ \wedge \forall c_1, c_2 \in columns(t_1) : name(c_1) \neq name(c_2) \\ \wedge \forall f_1, f_2 \in foreignKeys(t_1) : name(f_1) \neq name (f_2) \\ \wedge \forall f \in foreignKeys(t_1) : \exists t \in tables(d) \wedge name(t) = reference(f) \\ \wedge \forall r \in data(d): ( \exists t \in tables(d) : reference(r) = name(t) \\ \wedge \forall p \in pairs(r) :\\ \exists u \in columns(t) \cup foreignKeys(t) \cup primaryKey(t) : \\ reference(p) = name(u) \\ \wedge \forall e \in columns(t) \cup foreignKeys(t) \\ \wedge NOTNULL \in constrain(e) : \exists q \in pairs(r) \wedge name(e) = reference(q)) \\ \wedge \forall g \in columns(t) \cup foreignKeys(t) \\ \wedge UNIQUE \in constrain(g) : \\ \forall u_1, u_2 \in rows(t) : \\ value(pairByName(name(g), u_1) \neq value(pairByName(name(g), u_2) }{true}
 	\\ \\
 	false
 \end{cases}\\
& d \in Database
\end{align*}

@@

\paragraph{Mapping of Application} The goal of ORM is usually to determine how an instance is stored in the database. The ORM maps the structure of application classes to database schema. The assumption is a software is evolved continuously and evolutionary transformations produces consistent software, nevertheless to verify consistency of a software we need mapping between an existing application structure and database schema. 
\begin{align*}
&	\mathbf{\rho_{app}} \in Application \rightarrow Database\\
&	\rho_{app}(a) = \inference{state(a) = Consistent \; a}{\begin{gathered}
   db(\varnothing) \bullet addTable(classes(a)) \\ 
	\bullet (\forall c \in classes(a), p \in properties(c) : addProperty(name(c), p) \\
	\bullet (\forall c \in classes(a), e \in associations(c) : addAssociaiton(name(c), e) 
\end{gathered}}\\
& a \in Application
\end{align*}

@@

\paragraph{Mapping of Classes} The class is mapped into a table with  automatically created primary key. The primary key is always created with name Id and of the $\mathbb{N^{*}}$ type. The properties and associations has to be mapped separately.
\begin{align*}
&	\mathbf{\rho_{c}} : Class*  \rightarrow Table* \\
&	\rho_{c}(c) = \begin{cases}
		%\inference{c = \varnothing}{\varnothing}\\\\
		%\inference{|c| > 1}{\rho_c(head(c)) \cup \rho_c(tail(c))} \\ \\
		\inference{|c| = 1}{table(name(c), primaryKey("Id"), \varnothing, \varnothing)}  \end{cases}   \\ 	
& c \in Class
\end{align*}

@@

There can occur name collision during mapping in case there are more classes with property representing collection. To avoid that the mapping changes the name of the table representing a collection.
\begin{align*}
&	\mathbf{\rho_p} : Property* \rightarrow Column* \\ \\
&	\rho_{sp}(p) = \begin{cases}
		%\inference{p = \varnothing}{\varnothing}\\
		%\inference{|p| > 1}{\rho_{sp}(head(p)) \cup \rho_{sp}(tail(p))}\\\\
 		\inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 			\\ \wedge \exists t : t = table(name(owningClass(p)), d) 
		 }{\begin{gathered}
	  		column(name(p), \rho_t(type(p)), \varnothing )
		\end{gathered}} \\ \\
  		\inference{cardinality(p) = 1 \wedge mandatory(p)
			 \\ \wedge \exists t : t = table(name(owningClass(p)), d)
			\\ \wedge \not instanciated(owningClass(p))
		}{\begin{gathered}
 	 		column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
		\end{gathered}}
	\end{cases}\\
&	p \in Property
\end{align*}
\begin{align*}
&	\mathbf{\rho_p} : Property* \rightarrow Table* \\ \\
&	\rho_{cp}(p) = \begin{cases}
		%\inference{p = \varnothing}{\varnothing}\\
		%\inference{|p| > 1}{\rho_{cp}(head(p)) \cup \rho_{cp}(tail(p))}\\\\
     	\inference{cardinality(p) > 1 \wedge !\;mandatory(p) \\ 
     		\wedge 	\exists t : t = table(name(owningClass(p)), d)
     		}{\begin{gathered}
     			table(name(t)+name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), \varnothing), \\ foreignKey(name(p), 			name(owningClass(p)), NOTNULL) \\
    	\end{gathered}}\\ \\
     	\inference{ cardinality(p) > 1 \wedge mandatory(p) \\ 
     		\wedge \exists t : t = table(name(owningClass(p)), d) 
     		}{\begin{gathered}  
     			table(name(t)+name(p), primaryKey("Id"), \\ 
     			column("value", \rho_t(type(p)), NOTNULL), \\ 			foreignKey(name(p), name(owningClass(p)), NOTNULL)
    	 \end{gathered}}
 	\end{cases}\\
& 	p \in Property
\end{align*}

\paragraph{Mapping of Associations} An association is mapped into database as a foreign key in table representing associated class or as a table in case the association cardinality is one to many or many to many. Thus there are two mapping cases as for properties and the same solution is used to avoid name collisions.
\begin{align*}
&	\mathbf{\rho_a} : Association* \rightarrow ForeignKey* \\
&	\rho_a(a) = \begin{cases}
		%\inference{a = \varnothing}{\varnothing}\\
		%\inference{|a| > 1}{\rho_{a}(head(a)) \cup \rho_{a}(tail(a))}\\\\
		\inference{ startCardinality(a) = 0 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
		}{
			foreignKey(name(a), name(t),  \varnothing) 
	 	}
  \\ \\
 	 \inference{ startCardinality(a) = 1 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
 	 }{ 
		foreignKey(name(a), name(t),  NOTNULL)
	}
	 \end{cases}\\
&	 a \in Association
\end{align*}
\begin{align*} 
&	\mathbf{\rho_a} : Association* \rightarrow Table* \\
&	\rho_a(a) = \begin{cases}
 		%\inference{a = \varnothing}{\varnothing}\\
		%\inference{|a| > 1}{\rho_{a}(head(a)) \cup \rho_{a}(tail(a))}\\\\
		\inference{  startCardinality > 1 \\ \wedge \exists t \in tables(d) : name(owningClass(p)) = name(t) \\ \wedge \exists e \in tables(d) : name(reference(a)) = name(e)
  		}{\begin{gathered}  
		 table(name(t)+name(a), pk("Id"), \varnothing, \\ foreignKey(owningClass(a),owningClass(a), NOTNULL) \\ foreignkey(reference(a), reference(a), NOTNULL)) 
  		\end{gathered}}  
 	\end{cases}\\
&	 a \in Association
\end{align*}

@@

\paragraph{Add Table} The transformation creates a new table in the database. The new table should not contain any foreign keys.
\begin{align*}
&	\mathbf{addTable} : Table* \times Database \rightarrow Database \\ 
&	addTable(t, d) = \begin{cases}
		%\inference{|t| > 1}{d \bullet addTable(head(t)) \bullet addTable(tail(t))}
\\ \\	
		\inference{|t| = 1 \wedge \forall e \in tables(d) \wedge name(t) \neq name(e) \\ \wedge \forall f \in foreignKeys(t) : \\ s\exists e \in tables(d) \wedge name(e) = reference(f) }{tables(d) \cup t}
% \\ \\ 	Inconsistent \;d 
	 \end{cases}\\ 
&	t = \{x | x \in Table\}, d \in Database
\end{align*}

>>>>

\section{ORM Software}
\label{sec:evoIntro}
\begin{figure}
\centering
	\includegraphics[scale=0.3]{./images/system}
	\caption{The model of one generation of an ORM system consists of application persistent layer and relational database for storing instances.}
\label{fig:appStructure}
\end{figure}
A software implemented using an object-oriented language, which uses relational database as a data storage (as illustrated in the Figure \ref{fig:appStructure}) is object of interest of this paper. There are three important components of such a software: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM (we will use $\rho$ as a symbol for ORM), therefore we define a software as a triple consisting of an application and a database which are connected together by an ORM:
\begin{align*}
& \mathbf{software} = < Application, Database, \rho >\\
& \mathbf{sw} : Applications \times Databases \times ORM \rightarrow  Software \\ 
\\
& \mathbf{application} \in Software \rightarrow Application \\
& application(sw(a, d, \rho)) = a \\ \\
& \mathbf{database} \in Software \rightarrow Database \\
& database(sw(a, d, \rho)) = d \\\\
& a \in Application, d \in Database
\end{align*}

@@

The function querying the state (it means consistency) of a software is defined as:
\begin{align*}
& 	\mathbf{consistent} \in State \rightarrow Boolean \\
& 	consistent(state(s)) = \begin{cases}
 		\inference{s = Consistent \; s}{true}\\\\
 		\inference{s = Inconsistent \; s}{false}\\
 \end{cases}\\
& s \in Software
\end{align*}
The consistency of application and database is discussed later in this paper. The definition of software state only is provided:
\begin{align*}
&	\mathbf{state} : Software \rightarrow State \\
&	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \\ \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases} \\ 
& 	s \in Software
\end{align*}

@@

The application of a transformation on a software state (evolution) is marked by the $\oplus$ operator:
\begin{align*}
& \oplus \in State \times E \rightarrow State
\end{align*}
and similarly the $\circ$ and the $\bullet$ operators for evolving application and database:
\begin{align*}
& \circ \in State \times AppT \rightarrow State \\
& \bullet \in State \times DbT \rightarrow State
\end{align*}
The definition of the evolving operator is inspired by the monads (as used e.g. in Haskell programming language) so the state of the software can be passed from one evolution step to another, which helps us to concatenate evolution transformations.
\begin{align*}
& state(s) \oplus t = \begin{cases}
 	\inference{! consistent(state(s))}{Inconsistent \; s} \\ \\
 	\inference{consistent(state(s))}{state(sw(\Psi(s, t), \Phi(s,t))} 
\end{cases}\\ 
& s \in Software, t \in E
\end{align*}
\begin{align*}
& state(a) \circ e = \begin{cases}
 	\inference{!\;consistent(state(a))}{Inconsistent \; a} \\ \\
 	\inference{consistent(state(a))}{e(a)} 
\end{cases}\\ 
& a \in Application, e \in AppT
\end{align*}
\begin{align*}
&  state(d) \bullet e = \begin{cases}
 	\inference{!\;consistent(state(d))}{Inconsistent \; d} \\ \\
 	\inference{consistent(state(d))}{e(d)} 
\end{cases}\\ 
& d \in Database, e \in DbT
\end{align*}

@@

The definitions of all parts assume a set $Label$ exists which contains all possible identifiers and each modeled entity is unambiguously identified in the model. Next there is a set $Values$ containing all possible values, which can be used in the model.

A lot of introduced transformations can be applied on an ordered sets. The application proceeds on set consists in sequential application on set elements:
\begin{align*}
& 	f(x, y) = \begin{cases}
 		\inference{x = \varnothing}{y}\\
 		\inference{|x| > 1}{f(head(x), y) \cup f(tail(x), y}\\\\
 		\inference{|x| = 1}{effect}\\\\
 		\inference{otherwise}{Inconsistent y}
 \end{cases}\\\\
& y \in Software \vee y \in Application \vee y \in Database \\
& x \in Type* 
\end{align*}
where $x$ is a set of parameters. The effect of a transformation is provided only in next sections as the repetition of sequential application will be redundant. The $otherwise$ branch is consider implicit for all introduced transformations and thus it is not part of transformation definitions too.

@@

\paragraph{Foreign key} Foreign key is a reference to another table's primary key, it has an unique name and it can be constrained. The value of a foreign key is $\varnothing$ or a non-zero natural number.
\begin{align*}
&	\mathbf{ForeignKey} = < Label, Label, Constraint*> \\ 
&	\mathbf{fk} : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
&	\mathbf{name} : ForeignKey \rightarrow Label \\
&	name(fk(l, tRef, cons)) = lab  \\ \\
&	\mathbf{reference} : ForeignKey \rightarrow Label  \\
&	reference(fk(l, tRef, cons)) = tRef  \\ \\
&	\mathbf{constaints} : ForeignKey \rightarrow Constraint*  \\
&	constaints(fk(l, tRef, cons)) = cons  \\ \\
&	\mathbf{owningTable} : ForeignKey \times Database \rightarrow Table  \\
&	owningTable(fk, d) = \inference{\exists tab \in tables(d) \wedge fk \in foreignKeys(tab)}{tab}\\\\
&	l, tRef \in Label, cons  = \{ x | x \in Constraint\}, \\ 
& d \in Database, fk = \{ x | x \in ForeignKey\}
\end{align*}

@@

\paragraph{Stored Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete value of concrete column. Each row contains a name of a table it belongs to.
\begin{align*}
&	\mathbf{Row} = < Label, N, Pair* > \\
&	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
&	\mathbf{refTable} : Row \rightarrow Label \\
&	refTable(row(l, n, ps) = l \\ \\
&	\mathbf{pairs} : Row \rightarrow Pair* \\
&	pairs(row(l, n, ps)) = ps \\ \\
&	\mathbf{pkValue} : Row \rightarrow N \\
&	pkValue(row(l, n, ps)) = n \\ \\
&	 l \in Label, ps = \{ x | x \in Pair\}, n \in N
\end{align*}
The value stored in concrete column of the row is represented by a pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
&	\mathbf{Pair} = < Label, Value > \\
&	[Label \times Value] \rightarrow Pair\\ \\
&	\mathbf{reference} : Pair \rightarrow Label \\
&	reference([l,val]) = l \\ \\
&	\mathbf{value} : Pair \rightarrow Value \\ 
&	value([l,val]) = val \\ \\
&	l \in Label, val \in Value
\end{align*}
The definition of data allows definition of function $contains$ which verify if the table contains data in context of concrete database and $instantiated$, which search for instances of a class with given name and returns true if there are rows for the table in the database.
\begin{align*}
&	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
&	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
& l \in Label, ps  = \{ x | x \in Pair\}, t \in tables(d), d \in Database
\end{align*}
\begin{align*}
&	\mathbf{tableData} : Label \times Row* \rightarrow Row* \\
&	tableData(l, r) = \begin{cases}
 		%\inference{r = \varnothing}{\varnothing} \\
 		%\inference{|r| > 1}{tableData(head(r)) \cup tableData(tail(r))}\\ \\
&		\inference{|r| = 1 \wedge refTable(r) = l}{r}\\\\
& 		\inference{|r| = 1 \wedge refTable(r) \neq l}{\varnothing}
 \end{cases} \\ \\
&	r \in Row*, l \in Label
\end{align*}
\begin{align*}
&	\mathbf{instantiated} : Class \times Software \rightarrow Boolean \\
&	instantiated(c, s) = \begin{cases} \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  false
 \end{cases} \\ \\
&	 c \in classes(application(s)), s \in Software 
\end{align*}

\subsubsection{Database consistency}
The database consistency is founded on the consistency of structure and consistency between structure and stored values. The conditions for structure consistency are similar as for application structure. The consistency between structure and data is:  each row in a database  refers to a table in database schema and it has to have the structure corresponding to the one defined by a table, moreover the column and foreign key constrains has to be fulfilled  otherwise the database is inconsistent. The consistency of database is:
\begin{align*}
&	\mathbf{consistent} \in Database \rightarrow Boolean \\
&	state(d) = \begin{cases}
&		\inference{ \forall t_1, t_2 \in tables(d) : name(t_1) \neq name(t_2) \\ \wedge \forall c_1, c_2 \in columns(t_1) : name(c_1) \neq name(c_2) \\ \wedge \forall f_1, f_2 \in foreignKeys(t_1) : name(f_1) \neq name (f_2) \\ \wedge \forall f \in foreignKeys(t_1) : \exists t \in tables(d) \wedge name(t) = reference(f) \\ \wedge \forall r \in data(d): ( \exists t \in tables(d) : reference(r) = name(t) \\ \wedge \forall p \in pairs(r) :\\ \exists u \in columns(t) \cup foreignKeys(t) \cup primaryKey(t) : \\ reference(p) = name(u) \\ \wedge \forall e \in columns(t) \cup foreignKeys(t) \\ \wedge NOTNULL \in constrain(e) : \exists q \in pairs(r) \\ \wedge name(e) = reference(q)) \\ \wedge \forall g \in columns(t) \cup foreignKeys(t) \\ \wedge UNIQUE \in constrain(g) : \\ \forall u_1, u_2 \in rows(t) : \\ value(pairByName(name(g), u_1) \neq value(pairByName(name(g), u_2) }{true}
 	\\ \\
 	false
 \end{cases}\\
& d \in Database
\end{align*}

@@

\paragraph{Mapping of Application} The goal of ORM is usually to determine how an instance is stored in the database. The ORM maps the structure of application classes to database schema. The assumption is a software is evolved continuously and evolutionary transformations produces consistent software, nevertheless to verify consistency of a software we need mapping between an existing application structure and database schema. 
\begin{align*}
&	\mathbf{\rho_{app}} \in Application \rightarrow Database\\
&	\rho_{app}(a) = \inference{state(a) = Consistent \; a}{\begin{gathered}
   db(\varnothing) \bullet addTable(classes(a)) \\ 
	\bullet (\forall c \in classes(a), p \in properties(c) : \\ addProperty(name(c), p) \\
	\bullet (\forall c \in classes(a), e \in associations(c) : \\ addAssociaiton(name(c), e) 
\end{gathered}}\\
& a \in Application
\end{align*}

@@

\paragraph{Mapping of Classes} The class is mapped into a table with  automatically created primary key. The primary key is always created with name Id and of the $\mathbb{N^{*}}$ type. The properties and associations has to be mapped separately.
\begin{align*}
&	\mathbf{\rho_{c}} : Class*  \rightarrow Table* \\
&	\rho_{c}(c) = %\begin{cases}
		%\inference{c = \varnothing}{\varnothing}\\\\
		%\inference{|c| > 1}{\rho_c(head(c)) \cup \rho_c(tail(c))} \\ \\
		\inference{|c| = 1}{table(name(c), primaryKey("Id"), \varnothing, \varnothing)} % \end{cases} 
\\ 	
& c \in Class*
\end{align*}

@@

There can occur name collision during mapping in case there are more classes with property representing collection. To avoid that the mapping changes the name of the table representing a collection.
\begin{align*}
&	\mathbf{\rho_p} : Property* \rightarrow Column* \\ \\
&	\rho_{sp}(p) = \begin{cases}
		%\inference{p = \varnothing}{\varnothing}\\
		%\inference{|p| > 1}{\rho_{sp}(head(p)) \cup \rho_{sp}(tail(p))}\\\\
 		\inference{|p| = 1 \wedge cardinality(p) = 1 \wedge !\;mandatory(p)
 			\\ \wedge \exists t : t = table(name(owningClass(p)), d) 
		 }{\begin{gathered}
	  		column(name(p), \rho_t(type(p)), \varnothing )
		\end{gathered}} \\ \\
  		\inference{|p| = 1 \wedge cardinality(p) = 1 \wedge mandatory(p)
			 \\ \wedge \exists t : t = table(name(owningClass(p)), d)
			\\ \wedge \not instanciated(owningClass(p))
		}{\begin{gathered}
 	 		column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
		\end{gathered}}
	\end{cases}\\
&	p \in Property
\end{align*}
\begin{align*}
&	\mathbf{\rho_p} : Property* \rightarrow Table* \\ \\
&	\rho_{cp}(p) = \begin{cases}
		%\inference{p = \varnothing}{\varnothing}\\
		%\inference{|p| > 1}{\rho_{cp}(head(p)) \cup \rho_{cp}(tail(p))}\\\\
     	\inference{|p| = 1 \wedge cardinality(p) > 1 \wedge !\;mandatory(p) \\ 
     		\wedge 	\exists t : t = table(name(owningClass(p)), d)
     		}{\begin{gathered}
     			table(name(t)+name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), \varnothing), \\ foreignKey(name(p), 			name(owningClass(p)), NOTNULL) \\
    	\end{gathered}}\\ \\
     	\inference{|p| = 1 \wedge  cardinality(p) > 1 \wedge mandatory(p) \\ 
     		\wedge \exists t : t = table(name(owningClass(p)), d) 
     		}{\begin{gathered}  
     			table(name(t)+name(p), primaryKey("Id"), \\ 
     			column("value", \rho_t(type(p)), NOTNULL), \\ 			foreignKey(name(p), name(owningClass(p)), NOTNULL)
    	 \end{gathered}}
 	\end{cases}\\
& 	p \in Property
\end{align*}

\paragraph{Mapping of Associations} An association is mapped into database as a foreign key in table representing associated class or as a table in case the association cardinality is one to many or many to many. Thus there are two mapping cases as for properties and the same solution is used to avoid name collisions.
\begin{align*}
&	\mathbf{\rho_a} : Association* \rightarrow ForeignKey* \\
&	\rho_a(a) = \begin{cases}
		%\inference{a = \varnothing}{\varnothing}\\
		%\inference{|a| > 1}{\rho_{a}(head(a)) \cup \rho_{a}(tail(a))}\\\\
		\inference{|a| = 1 \wedge  startCardinality(a) = 0 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
		}{
			foreignKey(name(a), name(t),  \varnothing) 
	 	}
  \\ \\
 	 \inference{|a| = 1 \wedge  startCardinality(a) = 1 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
 	 }{ 
		foreignKey(name(a), name(t),  NOTNULL)
	}	 \end{cases}
\\
&	 a \in Association
\end{align*}
\begin{align*} 
&	\mathbf{\rho_a} : Association* \rightarrow Table* \\
&	\rho_a(a) =
% 	  \begin{cases}
 		%\inference{a = \varnothing}{\varnothing}\\
		%\inference{|a| > 1}{\rho_{a}(head(a)) \cup \rho_{a}(tail(a))}\\\\
		\inference{|a| = 1 \wedge   startCardinality > 1 \\ \wedge \exists t \in tables(d) : name(owningClass(p)) = name(t) \\ \wedge \exists e \in tables(d) : name(reference(a)) = name(e)
  		}{\begin{gathered}  
		 table(name(t)+name(a), pk("Id"), \varnothing, \\ foreignKey(owningClass(a),owningClass(a), NOTNULL) \\ foreignkey(reference(a), reference(a), NOTNULL)) 
  		\end{gathered}}  
 	%\end{cases}
\\
&	 a \in Association
\end{align*}

@@

\paragraph{Add Table} The transformation creates a new table in the database. The new table should not contain any foreign keys.
\begin{align*}
&	\mathbf{addTable} : Table* \times Database \rightarrow Database \\ 
&	addTable(t, d) =
		% \begin{cases}
		%\inference{|t| > 1}{d \bullet addTable(head(t)) \bullet addTable(tail(t))}
% \\ \\	
		\inference{|t| = 1 \wedge \forall e \in tables(d) \wedge name(t) \neq name(e) \\ \wedge \forall f \in foreignKeys(t) : \\ \exists e \in tables(d) \wedge name(e) = reference(f) }{tables(d) \cup t}
% \\ \\ 	Inconsistent \;d 
	 %\end{cases}
	 \\ 
&	t = \{x | x \in Table\}, d \in Database
\end{align*}
